---
output:
  github_document:
html_preview: true
---

<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->



```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
library(bcdata)
```

<a id="devex-badge" rel="Exploration" href="https://github.com/BCDevExchange/assets/blob/master/README.md"><img alt="Being designed and built, but in the lab. May change, disappear, or be buggy." style="border-width:0" src="https://assets.bcdevexchange.org/images/badges/exploration.svg" title="Being designed and built, but in the lab. May change, disappear, or be buggy." /></a>[![License](https://img.shields.io/badge/License-Apache%202.0-blue.svg)](https://opensource.org/licenses/Apache-2.0)[![Travis build status](https://travis-ci.org/bcgov/bcdata.svg?branch=master)](https://travis-ci.org/bcgov/bcdata)

# bcdata

An R package 📦 for searching & retrieving data from the [B.C. Data Catalogue]( https://catalogue.data.gov.bc.ca).

### Installation
You can install `bcdata` directly from this GitHub repository. To do so, you will need the [remotes](https://cran.r-project.org/package=remotes) package:

```{r, eval=FALSE}
install.packages("remotes")
```

Next, install and load the `bcdata` package using `remotes::install_github()`:

```{r, eval=FALSE}
remotes::install_github("bcgov/bcdata")
library(bcdata)
```



### Functions

- `bcdc_browse()` - Open the catalogue in your default browser
- `bcdc_search()` - Search records in the catalogue
- `bcdc_search_facet()` - List catalogue facet search options
- `bcdc_get_record()` - Print a catalogue record
- `bcdc_get_data()` - Get catalogue data
- `bcdc_query_geodata()` - Get catalogue geospatial data (data must be available as a [WMS/WFS service](https://www2.gov.bc.ca/gov/content?id=95D78D544B244F34B89223EF069DF74E))

### Usage

```{r, eval = TRUE, warnings=FALSE}
bcdc_search("airport")
```

```{r eval = TRUE}
bcdc_get_record("76b1b7a3-2112-4444-857a-afccf7b20da8")
```

### bcdc_get_data
To have a quick look at spatial data, you can simply print the object:
```{r}
airports <- bcdc_get_geodata("bc-airports", crs = 3857)
airports
```
If you need all the data, you need to retrieve the data using the `collect` function:
```{r}
airports %>% 
  collect()
```
This has the advantage of fine tuning your query without issuing a large request to the data catalogue. When you are finally ready to bring all the data you want into memory then you can use the `collect` function. This approach mimics that developed for `dbplyr`. You can read more about that [here](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html) 




### Project Status
This project is under active development.

### Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/bcdata/issues/).

### How to Contribute

If you would like to contribute to the package, please see our 
[CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.

### License

```
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the &quot;License&quot;);
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an &quot;AS IS&quot; BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
```

---
*This project was created using the [bcgovr](https://github.com/bcgov/bcgovr) package.* 
