---
output:
  github_document:
html_preview: true
---

<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->



```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
library(bcdata)
```

<a id="devex-badge" rel="Exploration" href="https://github.com/BCDevExchange/assets/blob/master/README.md"><img alt="Being designed and built, but in the lab. May change, disappear, or be buggy." style="border-width:0" src="https://assets.bcdevexchange.org/images/badges/exploration.svg" title="Being designed and built, but in the lab. May change, disappear, or be buggy." /></a>[![License](https://img.shields.io/badge/License-Apache%202.0-blue.svg)](https://opensource.org/licenses/Apache-2.0)[![Travis build status](https://travis-ci.org/bcgov/bcdata.svg?branch=master)](https://travis-ci.org/bcgov/bcdata)

# bcdata

An R package 📦 for searching & retrieving data from the [B.C. Data Catalogue]( https://catalogue.data.gov.bc.ca).

### Installation
You can install `bcdata` directly from this GitHub repository. To do so, you will need the [remotes](https://cran.r-project.org/package=remotes) package:

```{r, eval=FALSE}
install.packages("remotes")
```

Next, install and load the `bcdata` package using `remotes::install_github()`:

```{r, eval=FALSE}
remotes::install_github("bcgov/bcdata")
library(bcdata)
```



### Functions

- `bcdc_browse()` - Open the catalogue in your default browser
- `bcdc_search()` - Search records in the catalogue
- `bcdc_search_facet()` - List catalogue facet search options
- `bcdc_get_record()` - Print a catalogue record
- `bcdc_get_data()` - Get catalogue data
- `bcdc_get_geodata()` - Get catalogue geospatial data (data must be available as a [WMS/WFS service](https://www2.gov.bc.ca/gov/content?id=95D78D544B244F34B89223EF069DF74E))

### Usage

```{r, eval = FALSE, warnings=FALSE}
bcdc_search("airport")
```
```
Found 8 matches. Returning the first 100.
To see them all, rerun the search and set the 'n' argument to 8.
List of B.C. Data Catalogue Records

Number of records: 8
Titles:
1: Airport Capacity and Service Information
 ID: 553a8c7c-bca2-4b92-b925-ef6085724092
2: BC Airports
 ID: 76b1b7a3-2112-4444-857a-afccf7b20da8
3: Customs Ports of Entry
 ID: 4fac3ad6-8749-4741-ac98-527b23e4b0b2
4: BC Wildfire Active Weather Stations
 ID: c16867a3-a7ac-4c07-9a3e-0325e66e29c0
5: NTS BC Transport Points 1:250,000 - Digital Baseline Mapping (NTS)
 ID: 6ce711d4-5196-47fc-9bc0-0839b1aa2ca3
6: HelloBC Activities and Attractions Listing
 ID: 3a7dc21c-be34-4b33-adbe-fadd3aaba2d7
7: BC Climate Stations
 ID: d9c6cafe-0b24-4197-a91d-8448aa4d98c7
8: Sites Registry (Open Government License)
 ID: d5c6b8ed-c272-4c9e-8813-590a47b5c01c 

Access a single record by calling bcdc_get_record(ID)
with the ID from the desired record.
```
```{r eval = FALSE}
bcdc_get_record("76b1b7a3-2112-4444-857a-afccf7b20da8")
```
```
B.C. Data Catalogue Record:
    BC Airports 

Name: bc-airports (ID: 76b1b7a3-2112-4444-857a-afccf7b20da8 )
Permalink: https://catalogue.data.gov.bc.ca/dataset/76b1b7a3-2112-4444-857a-afccf7b20da8
Sector: Natural Resources
Licence: Open Government License - British Columbia
Type: Geographic 

Description:
     BC Airports identifies locations where aircraft may take-off and land. No guarantee is given that an identified point will be maintained to sufficient standards for landing and take-off of any/all aircraft.  It includes airports, aerodromes, water aerodromes, heliports, and airstrips. 

Resources: ( 4 )
  1: BC Airports - Custom Download
    description:  
    id: 03fdc03b-5487-4a2e-b7e3-a82e53db9ea1 
    format: other 
    access: Indirect Access 
    access_url: https://apps.gov.bc.ca/pub/dwds/addProductsFromExternalApplication.do?productTypeId=0&productId=17006 
  2: BC_Airports_Attribute_Values
    description: Valid attribute values for select attribute fields in BC Airports 
    id: fcccba36-b528-4731-8978-940b3cc04f69 
    format: xlsx 
    access: Direct Access 
    access_url: https://catalogue.data.gov.bc.ca/dataset/76b1b7a3-2112-4444-857a-afccf7b20da8/resource/fcccba36-b528-4731-8978-940b3cc04f69/download/wilmbvicgeobccriticalinfrastructuredocumentsbcairportsattributevalues.xlsx 
  3: WMS getCapabilities request
    description: For use in viewers such as ESRI tools use this truncated link: https://openmaps.gov.bc.ca/geo/pub/WHSE_IMAGERY_AND_BASE_MAPS.GSR_AIRPORTS_SVW/ows? [Click here for information on how to connect](http://www2.gov.bc.ca/gov/content/governments/about-the-bc-government/databc/geographic-data-and-services/map-services) 
    id: 4d0377d9-e8a1-429b-824f-0ce8f363512c 
    format: wms 
    access: Service 
    access_url: https://openmaps.gov.bc.ca/geo/pub/WHSE_IMAGERY_AND_BASE_MAPS.GSR_AIRPORTS_SVW/ows?service=WMS&request=GetCapabilities 
  4: KML Network Link
    description: For use in viewers such as Google Earth <BR> [Click here for information on how to connect](http://www2.gov.bc.ca/gov/content/data/geographic-data-services/web-based-mapping/map-services) 
    id: 5b9f22bf-eaa6-4640-ae65-9d3600672745 
    format: kml 
    access: Service 
    access_url: http://openmaps.gov.bc.ca/kml/geo/layers/WHSE_IMAGERY_AND_BASE_MAPS.GSR_AIRPORTS_SVW_loader.kml 

```
To have a quick look at data, you can simply print the object:
```{r}
airports <- bcdc_get_geodata("bc-airports", crs = 3857)
airports
```
If you need all the data, you need to retrieve the data using the `collect` function:
```{r}
airports %>% 
  collect()
```
This has the advantage of fine tuning your query without issuing a large request to the data catalogue. When you are finally ready to bring all the data you want into memory then you can use the `collect` function. This approach mimics that developed for `dbplyr`. You can read more about that [here](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html) 




### Project Status
This project is under active development.

### Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/bcdata/issues/).

### How to Contribute

If you would like to contribute to the package, please see our 
[CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.

### License

```
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the &quot;License&quot;);
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an &quot;AS IS&quot; BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
```

---
*This project was created using the [bcgovr](https://github.com/bcgov/bcgovr) package.* 
