---
title: "Get started with bcdata"
author: "" 
date: ""
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bcdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

library(bcdata)
```

The `bcdata` [R](https://www.r-project.org/) package contains functions for searching & retrieving data from the [B.C. Data Catalogue]( https://catalogue.data.gov.bc.ca).

The [B.C. Data Catalogue](https://www2.gov.bc.ca/gov/content?id=79B5224167334667A44C9E8B5143D0C5) is the place to find British Columbia Government data, applications and web services. Much of the data are released under the [Open Government Licence --- British Columbia](https://www2.gov.bc.ca/gov/content/data/open-data/open-government-licence-bc), as well as numerous other [licences](https://catalogue.data.gov.bc.ca/dataset?download_audience=Public).

You can install `bcdata` directly from GitHub using the [remotes](https://cran.r-project.org/package=remotes) package:

```{r, eval=FALSE}
install.packages("remotes")

remotes::install_github("bcgov/bcdata")
library(bcdata)
```

### `bcdc_browse()`

`bcdata::bcdc_browse()` let's you access the B.C. Data Catalogue home page directly from R---opening the web page in your default browser:

```{r, eval=FALSE}
## Take me to the B.C. Data Catalogue home page
bcdc_browse()
```

If you know the catalogue record human-readable _name_ or permanent _ID_ you can open the specific record page:

```{r, eval=FALSE}
## Take me to the B.C. Winery Locations catalogue record using the record name
bcdc_browse("bc-winery-locations") 

## Take me to the B.C. Winery Locations catalogue record using the record ID
bcdc_browse("1d21922b-ec4f-42e5-8f6b-bf320a286157")

## Take me to the catalogue search results for 'wineries'
bcdc_browse("wineries")
```


### `bcdc_search()`

`bcdata::bcdc_search()` let's you search the B.C. Data Catalogue. Let's search for catalogue records that contain "recycling" in the title:

```{r}
## Give me the catalogue search results for 'recycling'
bcdc_search("recycling")
```

You can set the number of records to be returned from the search and/or you can tailor your search using the catalogue search facets `license_id`, `download_audience`, `type`, `res_format`, `sector`, and `organization`:

```{r}
## Give me the first catalogue search result for 'recycling'
bcdc_search("recycling", n = 1)

## Give me the catalogue search results for 'recycling' where the data is tabular and the licence is Open Government Licence-British Columbia
bcdc_search("recycling", type = "Dataset", license_id = "2")
```

You can retrieve the _metadata_ for a single catalogue record by using the record _name_ or _ID_ with `bcdc_get_record()`. It is advised to use the permanent _ID_ rather than the human-readable _name_ in scripts or non-interactive situations to guard against future name changes of a record:

```{r}
## Give me the catalogue record metadata for `bc-first-tire-recycling-data-1991-2006`
bcdc_get_record("a29ad492-29a2-44b9-8693-d27a8cc8e686")
```

Finally, you can see all the valid values for the catalogue search facets using `bcdata::bcdc_search_facets()`:

```{r}
## Valid values for search facet 'type'
bcdc_search_facets(facet = "type")
```


### `bcdc_get_data()`

One you have located the B.C. Data Catalogue record with the data you want, you can use `bcdata::bcdc_get_data()` to download and read the data from the record.  Let's return to the B.C. Winery Locations data:

```{r}
## Get the data resource for the `bc-winery-locations` catalogue record
bcdc_get_data("bc-winery-locations") 

## OR use the permanent ID, likely better for scripts or non-interactive use 
bcdc_get_data("1d21922b-ec4f-42e5-8f6b-bf320a286157")
```
A catalogue record can have one or multiple data "resources"---as they are called. If there is only one resource it will be returned by default, as in the above `bc-winery-locations`. If there are multiple data resources you will need to specify which resource you want:

```{r}
## Get the data resource for the `bc-winery-locations` catalogue record
bcdc_get_data("bc-winery-locations") 

## OR use the permanent ID, likely better for scripts or non-interactive use 
bcdc_get_data("1d21922b-ec4f-42e5-8f6b-bf320a286157")
``

### `bcdc_query_geodata()`


### Further Reading
vignette("Querying spatial data with bcdata") demonstrates how to use `bcdc::query_geodata()` to fine tune a Web Map or Web Feature Service (WMS/WFS) request for geospatial data from the B.C. Data Catalogue.

