---
title: "Querying Spatial Data with bcdata"
author: "Sam Albers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# WORK IN PROGRESS

## Getting Started
First you need to load the package. We will also load the `sf` package to help us wokr with spatial data. You can learn more about the `sf` package [here](https://r-spatial.github.io/sf/):
```{r}
library(bcdata)
library(sf)
```


## Geospatial Data in the BC Data Catalogue
The BC Data Catalogue provides many datasets with spatial information over the web via a web feature service (wfs). Practically speaking, this means if we have an internet connection we can issue http requests to the wfs of the BC Data Catalogue and import the response data into R as an sf object very easily. The `bcdata` package provides a means to chose which layer you want and `dplyr` verbs to specifically tailor your request.  A `dbplyr` backend is implemented so that requests are executed lazily. This approach mimics the `dplyr` verb translation to `SQL` seen for many database types.

### Using dplyr verbs to fine tune your query
For the purposes of this example, let's consider [this](catalogue.data.gov.bc.ca/dataset/6a2fea1b-0cc4-4fc2-8017-eaf755d516da) layer in the catalogue. This layer is described here:
```{r}
bcdc_get_record("6a2fea1b-0cc4-4fc2-8017-eaf755d516da")
```

The key thing we are interested in here is that this is a *WFS record*. From this we know we can make use of `bcdc_query_geodata`.

```{r}
bcdc_query_geodata("6a2fea1b-0cc4-4fc2-8017-eaf755d516da")
```


### Schools in British Columbia
If we were interested in creating a quick map of most cities with a BC outline. First we would like to find the boundaries of BC layer. If I search for the "Province of British Columbia":

```{r}
bcdc_search("Province of British Columbia")
```
I can look and see that this entry:

> Province of British Columbia - Legally Defined Administrative Areas of BC
 ID: a7e32e45-63ae-4f5a-9275-9402b6deebdc
 
might be what I want. To look deeper into this record we can pass the ID to `bcdc_get_record`:
```{r}
bcdc_get_record("a7e32e45-63ae-4f5a-9275-9402b6deebdc")
```

This provides us considerable information on the data record including the fact that this record has wms capabilities and the name of the record. This is good news as we are easily ability to import this into R and create an sf object using the record name as:
```{r}
bc_bound <- bcdc_get_geodata("province-of-british-columbia-legally-defined-administrative-areas-of-bc")
```


And we use a simple plot to check if this indeed the layer we would like:

```{r}
plot(st_geometry(bc_bound))
```

With a BC outline, we can now try to extract information about cities in BC. We follow a similar workflow of `bcdc_search` -> `bcdc_get_record` -> `bcdc_get_geodata`:

```{r}
bcdc_search("cities")
```
This entry looks promising:
> BC Major Cities Points 1:2,000,000  (Digital Baseline Mapping)
 ID: b678c432-c5c1-4341-88db-0d6befa0c7f8

```{r}
bcdc_get_record("b678c432-c5c1-4341-88db-0d6befa0c7f8")
```
This again reveals wms capabilities so we can input that name:
```{r}
cities <- bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping")
```

Now we can combine our two layers into a quick map:
```{r}
plot(st_geometry(bc_bound))
plot(st_geometry(cities), add = TRUE, col = cities$POP_2000, pch = 19)
```

We can also make use of some further functionality offered by wfs by specifying conditions in `bcdc_get_geodata` in a similar way that would in `dplyr::filter()`. The advantage of using this approach is that focusses your query to the wfs meaning you only get exactly the data you ask for. For example, say we only wanted those cities classified as 'VILLAGE', we can add
`LONG_TYPE == 'VILLAGE'`. This gets translated in the background to a `SQL`-like statement and passed to the wfs service:
```{r}
bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping", 
                 LONG_TYPE == 'VILLAGE')
```
Note that because 'VILLAGE' is a character, it needs to be enclosed in quotes. If we wanted to chose multiple values of `LONG_TYPE` we can again write it as we normally would in R using the `%in%` operator:

```{r}
bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping", 
                 LONG_TYPE %in% c('CITY','VILLAGE'))
```

If you are a SQL/CQL user you might be more comfortable writing CQL queries 
directly rather than writing R expressions. You can do so by wrapping your query
as a character vector in the `CQL()` function:

```{r}
bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping", 
                 CQL("POP_2000 BETWEEN 10000 AND 20000"))
```

Note that this same result can be achieved using the `between()` function:

```{r}
bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping", 
                 between("POP_2000", 10000, 20000))
```

