---
title: "Getting Started with data"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# WORK IN PROGRESS

## Getting Started
First you need to loadthe package. We will also load the sf package to better enable our spatial data work:
```{r}
library(bcdata)
library(sf)
```


## Find your data set
There are several ways that you might arrive at the data set you want. The simplest way is to make use of the web-based BC Data Catalogue. The package provides a convenience function that allows you to quickly open a broswer to the BC Data Catalogue:
```{r,eval=FALSE}
bcdc_browse()
```
If you would like to further refine your search you can supply a string:
```{r}
bcdc_browse("cities")
```


## Geospatial Data
The BC Data Catalogue provides many dataset with spatial information via a web feature service (wfs). Practically speaking, this means that we can import this spatial data into R as an sf object very easily using the `bcdc_get_geodata` function. 


### Schools in British Columbia
If we were interested in creating a quick map of most cities with a BC outline. First we would like to find the boundaries of BC layer. If I search for the "Province of British Columbia":

```{r}
bcdc_search("Province of British Columbia")
```
I can look and see that this entry:

> Province of British Columbia - Legally Defined Administrative Areas of BC
 ID: a7e32e45-63ae-4f5a-9275-9402b6deebdc
 
might be what I want. To look deeper into this record we can pass the ID to `bcdc_get_record`:
```{r}
bcdc_get_record("a7e32e45-63ae-4f5a-9275-9402b6deebdc")
```

This provides us considerable information on the data record including the fact that this record has wms capabilities and the name of the record. This is good news as we are easily ability to import this into R and create an sf object using the record name as:
```{r}
bc_bound <- bcdc_get_geodata("province-of-british-columbia-legally-defined-administrative-areas-of-bc")
```
And we use a simple plot to check if this indeed the layer we would like:

```{r}
plot(st_geometry(bc_bound))
```

With a BC outline, we can now try to extract information about cities in BC. We follow a similar workflow of `bcdc_search` -> `bcdc_get_record` -> `bcdc_get_geodata`:

```{r}
bcdc_search("cities")
```
This entry looks promising:
> BC Major Cities Points 1:2,000,000  (Digital Baseline Mapping)
 ID: b678c432-c5c1-4341-88db-0d6befa0c7f8

```{r}
bcdc_get_record("b678c432-c5c1-4341-88db-0d6befa0c7f8")
```
This again reveals wms capabilities so we can input that name:
```{r}
cities <- bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping")
```

Now we can combine our two layers into a quick map:
```{r}
plot(st_geometry(bc_bound))
plot(st_geometry(cities), add = TRUE, col = cities$POP_2000, pch = 19)
```

We can also make use of some further functionality offered by wfs in the form of SQL-like strings supplied to the query parameter in `bcdc_get_geodata`. The advantage of using this approach is that focusses your query to the wfs meaning you only get exactly the data you ask for. For example, say we only wanted those cities classified as 'VILLAGE', we can add a small bit of SQL-like code:
```{r}
bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping", query = "LONG_TYPE='VILLAGE'")
```
Note that because 'VILLAGE' is a character, it needs to be enclosed in single quotes with the entire SQL statement enclosed in double quotes. If we wanted to chose multiple values of `LONG_TYPE` we can make use of the `IN` operator:

```{r}
bcdc_get_geodata("bc-major-cities-points-1-2-000-000-digital-baseline-mapping", query = "LONG_TYPE IN ('CITY','VILLAGE')")
```




