---
title: "An Exploration of Silviculture Data with bcdata"
author: "Hardy Griesbauer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore Silviculture Data with bcdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.height = 5,
  fig.width = 7
)

library(knitr)
library(bcdata)
library(dplyr)
library(ggplot2)
library(sf)
```


This vignette will demonstrate how you can use `bcdata` to access and explore British Columbia silviculture data.

Silviculture data for British Columbia are largely collected and stored through the [Reporting Silviculture Updates and Land Status Tracking System (RESULTS) database](https://www2.gov.bc.ca/gov/content?id=F62992DA5B324446AD5E1B5BFFA404CC)---including information on the management of openings, disturbances, silviculture activities and obligation declarations as required by the [Forest and Range Practices Act](https://www2.gov.bc.ca/gov/content?id=D65006C0326B4E0F838194044C10CA86). [RESULTS data sets are publically available through the B.C. Data Catalogue](https://catalogue.data.gov.bc.ca/dataset?q=RESULTS+silviculture) under the [Open Government Licence - British Columbia](https://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61).


## A Silviculture Data Exploration&mdash;Reforestation with Climate Considerations

[Western larch](https://en.wikipedia.org/wiki/Larix_occidentalis) has been identified as a tree species that will be well-adapted to projected future climates in northern British Columbia.  In 2013, forest policy was developed that allowed foresters to plant western larch outside of its natural range as a climate change adaptation silviculture measure. 

**How much western larch has been planted in the Prince George Natural Resource District?**  

Let's use the `bcdata` package to search, retrieve and explore the RESULTS silviculture data and answer this question. 


### Getting Started

To start, let's load the `bcdata` package. We will also load the `dplyr` and `ggplot2` packages to help us work with the data. You can learn more about the `dplyr` package [here](https://dplyr.tidyverse.org/) and the `ggplot2` package [here](https://ggplot2.tidyverse.org/):

```{r libs, eval = FALSE}
library(bcdata)
library(dplyr)
library(ggplot2)
```


### Getting the Data

We can gather the data we need to answer our question using the [RESULTS - silviculture forest cover dataset](https://catalogue.data.gov.bc.ca/dataset/258bb088-4113-47b1-b568-ce20bd64e3e3). First, let's take a look at the metadata record using `bcdc_get_record()`:

```{r get-record}
# Get the metadata using the human-readable record name
bcdc_get_record("results-forest-cover-silviculture")
```

We see that this is a [Web Service-enabled](https://www2.gov.bc.ca/gov/content?id=95D78D544B244F34B89223EF069DF74E) geospatial data set, so we can query and retrieve this geospatial data set using `bcdc_query_geodata()`:

```{r examine-silv-layer}
# Query the data using the permanent ID of the record to guard against name changes
bcdc_query_geodata("258bb088-4113-47b1-b568-ce20bd64e3e3")
```

This query shows that this data set has over 850,000 features and 159 fields.  Each feature is a treatment unit within a harvested opening, and contains information on the leading five tree species that are present in each treatment unit, including stems per hectare, age, and height.

Note that we have only _queried_ the data set so far---the data set would be too large (~1GB) to download efficiently.  So, let's use `filter` with `bcdc_query_geodata()` to refine our query _before_ we collect the data and import it into R.


### Refining a Geospatial Data Query


To address our question, we need the treatment data (1) from the Prince George Natural Resource District _and_ (2) that contain western larch.

First, we'll use the `bcdata` package to download the spatial boundary for the Prince George Natural Resource District:

```{r download-dpg}
## Create a spatial feature object named dpg
dpg <- bcdc_query_geodata("natural-resource-nr-district") %>%  
  filter(ORG_UNIT=="DPG") %>% # filter for Prince George Natural Resource District
  collect() # and collect the data
```

Let's plot this spatial object and double check we have we what we need:

```{r plot-dpg}
dpg %>% 
  ggplot() +
  geom_sf()
```

Now we have a spatial object that we can use as a bounding box to filter and download records in the RESULTS - silviculture layer from the Prince George Natural Resource District. 

We only need to download the treatments that have western larch planted. We can use the `bcdc_describe_feature()` helper function to examine the column names and types of the layer. In this case, we want to keep rows where the five `S_SPECIES_CODE_*` columns contain `"LW"`, the code for western larch.

```{r load-lw-data, cache = TRUE}
# Make a vector of tree species we are interested in 
# (in this case only LW for western larch)
spp_list = c("LW")
 
# Query and filter the data layer
trees_dpg <-
  bcdc_query_geodata("258bb088-4113-47b1-b568-ce20bd64e3e3") %>%
  filter(INTERSECTS(dpg)) %>% #filter for records that are within the DPG boundary
  filter(
      S_SPECIES_CODE_1 %in% spp_list |  #filter for LW records 
      S_SPECIES_CODE_2 %in% spp_list |
      S_SPECIES_CODE_3 %in% spp_list |
      S_SPECIES_CODE_4 %in% spp_list |
      S_SPECIES_CODE_5 %in% spp_list
  ) %>% 
  collect() #collect/download the data
```


### Exploring the Data

Let's look at the dimensions of this now much more manageable data object we have downloaded from the B.C. Data Catalogue:

```{r dim-data}
dim(trees_dpg)
```

We can see there are `r dim(trees_dpg)[1]` treatment units planted with western larch, and we can make a quick map of these harvested openings for the Prince George Natural Resource District:

```{r map-larch-plantations-dpg, message = FALSE}
trees_dpg %>% 
  ggplot() + 
  geom_sf() + 
  geom_sf(data = dpg, fill = NA) + #add the DPG spatial boundary
  theme_minimal()
```


We can also create some quick descriptive summaries of the data, treating the geospatial attribute table as a data frame in R, and answer our original question---how much western larch has been planted in the Prince George Natural Resource District?


#### What is the size and age distribution of larch plantations in the Prince George Natural Resource District in the year 2020?

```{r, warning = FALSE}
trees_dpg %>% 
  mutate(age = 2020 - REFERENCE_YEAR + S_SPECIES_AGE_1) %>% #create a plantation age column
  ggplot() +  #start a plot
  aes(x = age, y = FEATURE_AREA_SQM/10000) + #convert feature area to hectares
  geom_bar(stat = "sum") +
  scale_x_continuous(name = "Stand Age (years)",
                     limits = c(2, 30),
                     breaks = seq(from = 5, to = 30, by = 5)) +
  labs(y = "Sum Treatment Unit Area (ha)",
       title = "Area Planted with Western Larch by Stand Age\nin Prince George Natural Resource District",
       caption = paste0("Data sourced from the B.C. Data Catalogue\n on ",
                        Sys.Date(),
                        " using the bcdata R package")) +
  theme_minimal() +
  theme(legend.position = "none")
```

#### What is the Biogeoclimatic Ecosystem Classification (BEC) distribution of western larch plantations in the Prince George Natural Resource District?

We can download [British Columbia biogeoclimatic (BEC) data](https://catalogue.data.gov.bc.ca/dataset/bec-map) from the B.C. Data Catalogue using `bcdata` and join to our existing `trees_dpg` geospatial dataset using the `st_join` function from the `sf` package. You can learn more about `sf` [here](https://r-spatial.github.io/sf/).


```{r, cache = TRUE}
library(sf) #load the sf package

# Load the BEC data for Prince George Natural Resource District
bgc_dpg <- bcdc_query_geodata("WHSE_FOREST_VEGETATION.BEC_BIOGEOCLIMATIC_POLY") %>%
  filter(BBOX(dpg, crs = `EPSG:3005`)) %>% #note filtering with a BoundingBox
  collect()

# Join the BEC data with our tress_dpg geospatial data set
trees_bec_dpg <- trees_dpg %>% 
  st_join(bgc_dpg[, "MAP_LABEL"]) #join BEC data for each polygon
```

Now, we can summarize the area planted with western larch by biogeoclimatic unit:

```{r, warning = FALSE}
trees_bec_dpg %>% 
  group_by(MAP_LABEL) %>% # group polygons by biogeoclimatic unit
  summarise(Area = sum(FEATURE_AREA_SQM)/10000) %>%
  ggplot() +
  aes(x = MAP_LABEL, y = Area) +
  geom_col() +
  labs(y = "Sum Treatment Unit Area (ha)",
       x = "Biogeoclimatic Unit",
       title = "Area Planted with Western Larch by Biogeoclimatic Unit\nin Prince George Natural Resource District",
       caption = paste0("Data sourced from the B.C. Data Catalogue\n on ",
                        Sys.Date(),
                        " using the bcdata R package")) +
  theme_minimal() +
  theme(legend.position = "none")
```
