---
title: 'bcdata: An R package for searching & retrieving data from the B.C. Data Catalogue'
authors:
- affiliation: 1
  name: Andy Teucher
  orcid: 0000-0002-7840-692X
- affiliation: 2
  name: Sam Albers
  orcid: 0000-0002-9270-7884
- affiliation: 2
  name: Stephanie Hazlitt
  orcid: 0000-0002-3161-2304
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
bibliography: paper.bib
tags:
- R
- open data
- WFS
- British Columbia
affiliations:
  name: State of Environment Reporting, Ministry of Environment and Climate Change
    Strategy, British Columbia Provincial Government
  index: 1
---


```{r options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, messages = FALSE, fig.width = 8, fig.height = 12, fig.path = "")
```

```{r packages, warning=FALSE, message=FALSE, echo = FALSE}
library(bcdata)
library(sf)
library(ggplot2)
```

# Introduction
The British Columbia government hosts over 2000 tabular and spatial data sets in the B.C. Data Catalogue (@bcdc).  Most provincial spatial data is available through the B.C. Data Catalogue under an open licence, via a Geoserver Web Feature Service (WFS). WFS is a powerful and flexible service for distributing geographic features over the web, that supports both spatial and non-spatial querying.  The bcdata package for the R programming language (@RCore) wraps this functionality and enables R users to efficiently query and directly read spatial data from the B.C. Data Catalogue into their R session. The bcdata package implements a novel application of dbplyr using a web service backend, where a locally constructed query is processed by a remote server. The data is only downloaded, and loaded into R as an ‘sf’ object, once the query is complete and the user requests the final result. This allows for fast and efficient spatial data retrieval using familiar dplyr syntax. The package also provides functionality that enables users to search and retrieve many other types of data and metadata from the B.C. Data Catalogue, thereby connecting British Columbia open data holdings with the vast capabilities of R.

# Usage 

bcdata connects to the B.C. Data Catalogue through a few key functions:

- `bcdc_browse()` - Open the catalogue in your default browser
- `bcdc_search()` - Search records in the catalogue
- `bcdc_search_facets()` - List catalogue facet search options
- `bcdc_get_record()` - Print a catalogue record
- `bcdc_tidy_resources()` - Get a data frame of resources for a record
- `bcdc_get_data()` - Get catalogue data
- `bcdc_query_geodata()` - Get & query catalogue geospatial data available through a [Web Service](https://www2.gov.bc.ca/gov/content?id=95D78D544B244F34B89223EF069DF74E)

### `bcdc_get_data()`

Once you have located the B.C. Data Catalogue record with the data you want, you can use `bcdata::bcdc_get_data()` to download and read the data from the record.  You can use the record name, permanent ID or the result from `bcdc_get_record()`. Let's look at the B.C. Highway Web Cameras data:

```{r eval = FALSE}
## Get the data resource for the `bc-highway-cams` catalogue record
bcdc_get_data("bc-highway-cams")

## OR use the permanent ID, which is better for scripts or non-interactive use
bcdc_get_data("6b39a910-6c77-476f-ac96-7b4f18849b1c")

## OR use the result from bcdc_get_record()
my_record <- bcdc_get_record("6b39a910-6c77-476f-ac96-7b4f18849b1c")
bcdc_get_data(my_record)
```

A catalogue record can have one or multiple data files---or "resources". If there is only one resource, `bcdc_get_data()` will return that resource by default, as in the above `bc-highway-cams` example. If there are multiple data resources you will need to specify which resource you want.

### `bcdc_query_geodata()`

While `bcdc_get_data()` will retrieve the geospatial data for you, sometimes the geospatial file is very large---and slow to download---and/or you may only want _some_ of the data. `bcdc_query_geodata()` let's you query catalogue geospatial data available as a Web Service using `select` and `filter` functions (just like in [`dplyr`](https://dplyr.tidyverse.org/). The `bcdc::collect()` function returns the `bcdc_query_geodata()` query results as an [`sf` object](https://r-spatial.github.io/sf/) in your R session.

Let's get the Capital Regional District boundary from the [B.C. Regional Districts geospatial data](https://catalogue.data.gov.bc.ca/dataset/d1aff64e-dbfe-45a6-af97-582b7f6418b9)---the whole file takes 30-60 seconds to download and I only need the one polygon, so why not save some time:

```{r regional_districts, fig.height = 5, fig.width = 7}
## Find the B.C. Regional Districts catalogue record
bcdc_search("regional districts administrative areas", res_format = "wms", n = 1)

## Get the metadata for the B.C. Regional Districts catalogue record
bc_regional_districts_metadata <- bcdc_get_record("d1aff64e-dbfe-45a6-af97-582b7f6418b9")

## Have a quick look at the geospatial columns to help with filter or select
bcdc_describe_feature(bc_regional_districts_metadata)

## Get the Capital Regional District polygon from the B.C. Regional
## Districts geospatial data
my_regional_district <- bcdc_query_geodata(bc_regional_districts_metadata) %>%
  filter(ADMIN_AREA_NAME == "Capital Regional District") %>%
  collect()

## Plot the Capital Regional District polygon with ggplot()
my_regional_district  %>%
  ggplot() +
  geom_sf() +
  theme_minimal()
```

# Conclusion

Connecting the R programming language with the B.C. Data Catalogue creates an API that allows for the usage of cutting edge statistical and plotting capabilities with a vast collection of open and public data. The enables usages in a modern data science context and provides a pathway to generate insights from public data. 

# Acknowledgements
Author order was determined randomly using the following R code: `set.seed(42); sample(c("Teucher","Hazlitt","Albers"), 3)`

# References
